Contracts for unions and intersections are hard for Nickel

### 2021-05-31 paper.tex
Make the syntax of examples uniform + fix typos

\lstdefinelanguage{nickel}{
    keywords=[1]{
      if,
      then,
      else,
      switch,
    },
    keywords=[2]{
      let,
      rec,
      fun,
      in
    },
    keywords=[3]{
      Num,
      Bool,
      Str,
      List,
      Dyn,
      forall,
    },
    keywordsprefix=\#,
    keywords=[4]{
      true,
      false,
      null
    },
    keywords=[5]{
      doc,
      default,
    },
    sensitive=true, % keywords are case-sensitive
    morecomment=[l]{//}, % l is for line comment
    morecomment=**[is][\color{gray}]{\$}{\$},
    morestring=[b]", % defines that strings are enclosed in double quotes
    morestring=[s]{m\#"}{"\#m},
    moredelim=[is][\color{gray}]{\$}{\$},
    literate=
        *{->}{$\rightarrow$}1
        {...}{\ldots}1
        {=>}{$\Rightarrow$}1
        {>=}{$\geq$}1
        {<=}{$\leq$}1
} %

@@

We will use the syntax of Nickel to write most examples throughout this paper.
Syntactically, Nickel looks like a simple ML language with records. \nickel{let
id = value in exp} is used to bind a value to an identifier in an expression.
Functions are declared using the \nickel{fun arg1 ... argn => body} syntax. List
are written as \nickel{[x1, ..., xn]} and records as \nickel{\{ field1 = value1,
..., fieldn = valuen \}}. Some examples uses constructions that are not
currently available in Nickel for illustration purpose: in particular, we will
use \nickel{A & B} (resp.  \nickel{A | B}) for the intersection (resp. union) of
contracts (or types) \nickel{A} and \nickel{B}. Finally, we will write
\nickel{exp @ C}\footnote{The current Nickel implementation actually uses the
syntax \nickel{exp | C} for contract application, but we rather keep \nickel{|}
for unions here}  for the application of a contract \nickel{C} to the expression
\nickel{exp}.

@@

Assertions are usually mainly used for correctness by enforcing pre- and
post-conditions of functions at runtime. In their foundational
paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
principled approach to run-time assertion checking called contracts. As opposed
to plain assertions, contracts can be composed naturally into more complex ones.
Nickel features such a contract system. For example, contracts nicely support
higher-order functions. If the user has written an \nickel{Even} contract for
numbers, they can use it right away to form a function contract \nickel{#Even ->
#Int} contract, as illustrated in Figure~\ref{fig:contract-higher-order}. This
contract will check at each call that the argument is an even number
(pre-condition) and that the return value is a integer (post-condition). The
\nickel{#} character distinguishes contracts in Nickel. Contracts also introduce
the notion of blame which is crucial to good error reporting in presence of
higher-order contracts and other composed contracts.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Even = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 2 == 0) in
let Int = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 1 == 0) in
{divBy2 | #Even -> #Int = fun x => x/2}
\end{lstlisting}
  \end{center}
\caption{Higher-order contract}
\label{fig:contract-higher-order}
\end{figure}

@@

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nixos.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\end{figure}

Contracts have another perk: they correspond to the type casts of gradual
typing, a discipline for mixing static and dynamic typing in the same language.
As such, they are a precious ingredient for handling the interaction between 
statically typed and dynamically typed code. While we focused on the writing of
configuration, Nickel is a proper programming language that is also designed to
write generic and reusable library code. There, the typing trade-off is
different, and static typing is more adapted. Nickel has a gradual type system,
which is yet another argument for having contracts. Let us end this section by
stressing the following points:
\begin{itemize}
    \item \textbf{Even in a purely dynamically typed language, contracts are an
        appealing feature for validation and correctness}. The gradual typing
        part enters the general picture but is out of the scope of this paper.
    \item \textbf{User-defined contracts are essential for validation}. Some
        contract systems (as the one of~\cite{RootCauseOfBlame}, that will be
        extensively discussed later in the paper) do not feature user-defined
        contracts.  Rather, a fixed set of built-in contracts such as
        \nickel{Number} or \nickel{String} are provided, very much like base
        types. The absence of user-defined contracts is understandable when
        using contracts as a dynamic typing discipline for correctness. But a
        system without user-defined contracts is limited in the same way as
        static typing with respect to validation.
\end{itemize}

@@

% \begin{figure}[h]
% \begin{center}
% \begin{lstlisting}[language=nickel]
% let f = (fun n => n * 2) @ (Integer -> even)
% \end{lstlisting}
%   \end{center}
% \caption{A function wrapped with a contract stating that it always returns
% an even number}
% \label{fig:num-to-even}
% \end{figure}

@@

The soundness of all these optimizations is tied to the invariance of the
semantics of programs with respect to specific substitutions. Inlining requires
that one can replace the application of a function by its body, which is
basically $\beta$-reduction: as long as the arguments are evaluated following
the language's strategy, it is a valid transformation. On the other hand,
CSE-like transformations require an expansion property of the form

$$M[N] \simeq let~x~=~N~in~M[x]$$

This equation clearly fails in a call-by-value language with side effects:

\begin{lstlisting}[language=Nickel]
let f x = print "hi";(x+1) in (f 1,f 1)
(*@ $\not \simeq$ @*) let f x = print "hi";(x+1) in
(*@ \phantom{$\not \simeq$} @*) let y = f 1 in (y,y)
\end{lstlisting}

In the example above, the first term prints \nickel{"hi"} two times while the
second term prints it only once. However, this property does hold for
\emph{pure} terms, that are terms without side-effects. In a pure language, or a
language with effect tracking like Haskell\todo{Cite a different language, or
remove citation}, pure terms can be safely identified for applying
optimizations.

Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of unions and
intersections can make this equivalence \todo{name it} unsound \emph{even in a
pure setting}, making the optimization of
Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.

@@

\change{These next few labels need to be inside figures,
they are not working}
\label{fig:union-adt-ex}
\begin{lstlisting}[title={Tree-like data structure}]
type BinaryTree =
Integer | (BinaryTree, BinaryTree)
\end{lstlisting}

In addition to internal program representations, unions are also useful to model
external data:
\begin{itemize}
    \item To represent nullable values(\ref{fig:union-nullable-ex}).
    \item For plain enumerations (\ref{fig:union-enum-ex}).
    \item To support different representations of the same data
        (\ref{fig:union-alt-ex}).
\end{itemize}
All of which are common in configurations.

\label{fig:union-nullable-ex}
\begin{lstlisting}[title={Nullable values}]
optionalValue : String | Null
\end{lstlisting}

\label{fig:union-enum-ex}
\begin{lstlisting}[title={Plain enumeration}]
protocol : `Http | `Ftp | `Sftp
\end{lstlisting}

\label{fig:union-alt-ex}
\begin{lstlisting}[title={Alternative representations}]
person : {name : String, age : Int}
  | {name : String, dateOfBirth : String}
\end{lstlisting}

@@

\label{fig:intersection-subytping-ex}
\begin{lstlisting}[title={Value subtyping}]
1 : Int & Float
\end{lstlisting}

@@

\label{fig:intersection-overloading-ex}
\begin{lstlisting}[title={Overloading}]
incr: (Int -> Int) & (Float -> Float) =
  fun x => x + 1
\end{lstlisting}

@@

\label{fig:intersection-mixins-ex}
\begin{lstlisting}[title={Structural mixins}]
Comparator: {cmp: T -> T -> Bool}
Shower: {show: T -> String}

@@

\label{fig:intersection-dual}
\begin{lstlisting}[title={Intersection as the dual of unions}]
f : (Num | String) -> Bool
(*@ $\simeq$ @*) f : (Num -> Bool) & (String -> Bool)
\end{lstlisting}

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = (fun x => x - 7) @ Positive -> Positive in
(f 10) + (f 5)
\end{lstlisting}
\caption{Wrong positive to positive function}
\label{fig:pos-to-pos}
\end{figure}

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = (fun x => x - 7) @ (Positive -> Positive) | (Positive -> NonPositive) in
(f 10) + (f 5)
\end{lstlisting}
\caption{Still wrong positive to positive or positive to non-positive function}
\label{fig:wrong-union-function}
\end{figure}

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Inlined]
let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
(f 1 true, f 1 false)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
let f = (fun x y => if y then x else "False") @ Number -> (Boolean -> Number) | (Boolean -> String) in
let g x = f 1 x in
(g true, g false)
\end{lstlisting}
\caption{Equivalent programs, with inlining or CSE applied}
\label{fig:optimized-programs}
\end{figure}

Unexpectedly, these two programs behave very differently, the first one will return
\nickel{(1, "False")}, without failing.
While the second one will fail, since the single application of \nickel{f 1},
applied to both \nickel{true} and \nickel{false} will be able to find the error
on the union contract (\nickel{(Boolean -> Number) | (Boolean -> String)}).

@@

Summing it up, wrapping a value with a union of higer order contracts means that
that value now has a state, that may get modified silently at every use site.
Both reviewed papers suffer from this defect since it is the semantics
of unions that interact badly with code manipulation operations,
not a particular implementation.
This addition of global state complicates both the work of the language
implementor, as well as the language user, since it adds non trivial
side effects to your code.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Implementation]
x @ A & B
(*@ $\simeq$ @*)
(x @ A) @ B
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Problem]
let g = (fun x => x) @ (Number -> Number) & (String -> String) in
g 1
\end{lstlisting}
\caption{Implementations of intersection contracts: first iteration}
\label{fig:inter-contracts}
\end{figure}

Not all hope is lost, we could pass some extra shared state
to each one of the sub contracts and let them only fail if the other
side of the contract has failed as well, as shown in Figure
\ref{fig:inter-contracts-2}, where the shared state is
represented by a label \nickel{l}.
This idea is still quite reasonable, for instance, when executing
\nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
that gets checked is the argument of \nickel{String -> String} over the function parameter, so
something like \nickel{1 @ String}.
Of course, this fails, but before raising a violation, this contract could
check if the other side of the intersection (\nickel{Number -> Number}) has
also tried to raise negative blame on the same parameter ()\nickel{1}), since
it hasn't, the contract would return normally, and would mark itself as failed.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Implementation]
x @ A & B
(*@ $\simeq$ @*)
(x @ A[l]) @ B[l]
\end{lstlisting}
\caption{Implementations of intersection contracts: second iteration}
\label{fig:inter-contracts-2}
\end{figure}

On the other hand, if you were to do something like
\nickel{g true}, that would fail also negatively the contract
\nickel{String -> String} and, when checking if \nickel{true}
is a correct parameter for \nickel{Number -> Number} it would check
if the other side of the intersection has failed, and since it
has, a violation would be raised and the whole program would
correctly fail.

This idea, which is an oversimplification of the ideas
presented by Williams, Morris, and Wadler, seems to be spot on.
However, it presents a major problem, it's not compatible with flat contracts,
as presented in Section \ref{sec:flat-contracts}.
Consider the example presented on Figure \ref{fig:inter-flat-contracts},
if the implementation discussed previously were to be used, we may end up with
something like this \nickel{((fun x => x) @ (String -> String)[l]) @ C[l]}, where
\nickel{l} represents the shared state.
At this point, the intuitive thing to do is to check contract \nickel{C}, by checking
if \nickel{((fun x => x) @ (String -> String)[l]) 0 == 0}.
This will fail, since applying \nickel{0} to a \nickel{String -> String} contract
fails negatively.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let C = fun f => (f 0) == 0 in
let g = (fun x => x) @ (String -> String) & C
\end{lstlisting}
\caption{Intersection and flat contracts}
\label{fig:inter-flat-contracts}
\end{figure}

@@

These expectations are not really met by how unions and intersections
contracts are supposed to work.
For example, consider the Nickel code shown on Figure
\ref{fig:intersection-distrbution}, at first sight, this code should
complete execution succesfully, returning \nickel{(1, 1)},
however, under both the Keil and Thiemann, as well as the
Williams, Morris, and Wadler works, this program will fail, with a violation
of the contract imposed to \nickel{g}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = fun x y => x in
let g = f @ (Number -> Number -> Number) & (Number -> Boolean -> Number) in
let h = g 1 in
(h 1, h true)
\end{lstlisting}
\caption{Intersection contracts don't distribute}
\label{fig:intersection-distrbution}
\end{figure}

The reason this happens is that
there is a fundamental property of how intersection
contracts behave: the choice of which branch to choose is made exactly
when the intersection is opened up, and not delayed as much as possible,
as one would like.
This means, that the contract presented on Figure \ref{fig:intersection-distrbution},
\nickel{(Number -> Number -> Number) & (Number -> Boolean -> Number)},
will not behave the same way that
\nickel{(Number -> (Number -> Number) & (Boolean -> Number)},
when, with an intuitive view of how intersections should behave,
these two expressions should be logically equivalent.

It is worth noting, that intersection types do have this property,
since subtyping would allow to correctly type check a term like
\nickel{g 1} as both \nickel{Number -> Number} as well as
\nickel{Boolean -> Number}, and then deducing that \nickel{g 1}
can be given the type \nickel{(Number -> Number) & (Boolean -> Number)}.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
x @ A & (B | C)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Transformed]
x @ (A & B) | (A & C)
\end{lstlisting}
\caption{Contract transformation performed by the Keil and
Thiemann work}
\label{fig:kt-contract-transf}
\end{figure}

This straightforward idea makes composing the Keil and Thiemann
solution with other existing systems more difficult; in the words
of Williams, Morris, and Wadler (see Section~\ref{sec:will-morr-wadl}):
``(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
special cases can hinder composition.''\cite{RootCauseOfBlame}.
\info{(Yann) Does
it also make it less efficient? If $(A \& B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
And it also could harm performance, or at least complicate an implementation,
since some sharing mechanism should be used to avoid checking the same contract,
over the same value, multiple times, exemplified by contract \nickel{A} on
Figure \ref{fig:kt-contract-transf}, that gets duplicated.
\unsure{Arnaud says: this whole
section is too abstract give concrete code and explain what goes
wrong. It's fine to give code in Nickel syntax}

@@

As an example, consider the programs shown on Figure
\ref{fig:kt-non-deterministic}, if the Original program
gets executed, it will spawn two instances of said program,
one with the first branch of the intersection (\nickel{3 @ Number}),
and the other one with the second branch (\nickel{3 @ Positive}),
and these will get joined once they reach an equal point, in this case
it will be \nickel{3}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
3 @ Number & Positive
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=First Program]
3 @ Number
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Second Program]
3 @ Positive
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Joined Program]
3
\end{lstlisting}
\caption{Non deterministic intersection contract opening}
\label{fig:kt-non-deterministic}
\end{figure}

The second set of rules is deterministic, and it is intended
to be simpler to implement.
However, this second set needs to perform operations
on the evaluation context, to decide whether or not to evaluate
a given contract.
\footnote{You can check more on this topic on Section 5.1 of
\cite{KeilThiemannUnionIntersection}}
For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
when using the value \nickel{(fun x => x) @ (String -> String)}
inside the execution of contract \nickel{C}, and applying it to \nickel{0},
the contract runtime would check the context of execution and it would
deduce that contract \nickel{String -> String} does not apply in this case,
skipping it.

@@

Altough this is true, it's a weaker statement of what we'd like to have,
consider the two contexts presented on Figure \ref{fig:valid-contexts},
the first one is a valid context for a \nickel{Number -> Number} function,
that is, you could replace the $\square$ by some function wrapped in a
\nickel{Number -> Number} contract, and the context will not violate it.
Similarly, the second context from the same figure is a valid
\nickel{Boolean -> Boolean} context.

>>>>

\lstdefinelanguage{nickel}{
    keywords=[1]{
      if,
      then,
      else,
      switch,
    },
    keywords=[2]{
      let,
      rec,
      fun,
      in
    },
    keywords=[3]{
      forall,
    },
    keywordsprefix=\#,
    keywords=[4]{
      true,
      false,
      null
    },
    keywords=[5]{
      doc,
      default,
    },
    sensitive=true, % keywords are case-sensitive
    morecomment=[l]{//}, % l is for line comment
    morecomment=**[is][\color{gray}]{\$}{\$},
    morestring=[b]", % defines that strings are enclosed in double quotes
    morestring=[s]{m\#"}{"\#m},
    moredelim=[is][\color{gray}]{\$}{\$},
    literate=
        *{->}{$\rightarrow$}1
        {...}{\ldots}1
        {@&}{$\cap$}1
        {@|}{$\cup$}1
        {=>}{$\Rightarrow$}1
        {>=}{$\geq$}1
        {<=}{$\leq$}1
} %

@@

We will use the syntax of Nickel to write most examples throughout this paper.
Syntactically, Nickel looks like a simple ML language with records. \nickel{let
id = value in exp} is used to bind a value to an identifier in an expression.
Functions are declared using the \nickel{fun arg1 ... argn => body} syntax. List
are written as \nickel{[x1, ..., xn]} and records as \nickel{\{field1 = value1,
..., fieldn = valuen\}}. The application of a contract \nickel{C} to the
expression \nickel{exp} is written \nickel{exp | C}\footnote{The current
    implementation requires a \nickel{#} character in front of custom contracts,
    whose application would rather be \nickel{exp | #C}, but it is irrelevant
and subject to change in the future.}. We will sometimes attach
contracts directly to an identifier instead of the corresponding bound
expression, that is: \nickel{let id | C = value in exp} is syntactic sugar for
\nickel{let id = (value | C) in exp}.

Some examples resort to constructions that are not currently available in Nickel
for illustration purpose: we will use \nickel{A @& B} (resp. \nickel{A @| B})
for the intersection (resp. union) of contracts (or types) \nickel{A} and
\nickel{B}. \nickel{@|} and \nickel{@&} bind more loosely than arrows, meaning
that \nickel{A -> B @& C -> D} is to be parsed as \nickel{(A -> B) @& (C -> D)}.

@@

Assertions are usually mainly used for correctness by enforcing pre- and
post-conditions of functions at runtime. In their foundational
paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
principled approach to run-time assertion checking called contracts. As opposed
to plain assertions, contracts can be composed naturally into more complex ones.
Nickel features such a contract system. For example, contracts nicely support
higher-order functions. If the user has written an \nickel{Even} contract for
numbers, they can use it right away to form a function contract \nickel{Even ->
Int} contract, as illustrated in Figure~\ref{fig:contract-higher-order}. This
contract will check at each call that the argument is an even number
(pre-condition) and that the return value is a integer (post-condition). The
\nickel{#} character distinguishes contracts in Nickel. Contracts also introduce
the notion of blame which is crucial to good error reporting in presence of
higher-order contracts and other composed contracts.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Even = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 2 == 0) in
let Int = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 1 == 0) in
{divBy2 | Even -> Int = fun x => x/2}
\end{lstlisting}
  \end{center}
\caption{Higher-order contract}
\label{fig:contract-higher-order}
\end{figure}

@@

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | Host,
  port | Port,
} in
{host = "nixos.org", port = 80}
| Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\end{figure}

Contracts have another perk: they correspond to the type casts of gradual
typing, a discipline for mixing static and dynamic typing in the same language.
As such, they are a precious ingredient for handling the interaction between 
statically typed and dynamically typed code. While we focused on the writing of
configuration, Nickel is a proper programming language that is also designed to
write generic and reusable library code. There, the typing trade-off is
different, and static typing is more adapted. Nickel has a gradual type system,
which is yet another argument for having contracts. Let us end this section by
stressing the following points:
\begin{itemize}
    \item \textbf{Even in a purely dynamically typed language, contracts are an
        appealing feature for validation and correctness}. The gradual typing
        part enters the general picture but is out of the scope of this paper.
    \item \textbf{User-defined contracts are essential for validation}. Some
        contract systems (as the one of~\cite{RootCauseOfBlame}, that will be
        extensively discussed later in the paper) do not feature user-defined
        contracts.  Rather, a fixed set of built-in contracts such as
        \nickel{Num} or \nickel{Str} are provided, very much like base
        types. The absence of user-defined contracts is understandable when
        using contracts as a dynamic typing discipline for correctness. But a
        system without user-defined contracts is limited in the same way as
        static typing with respect to validation.
\end{itemize}

@@

% \begin{figure}[h]
% \begin{center}
% \begin{lstlisting}[language=nickel]
% let f = fun n => n * 2 @ (Integer -> even)
% \end{lstlisting}
%   \end{center}
% \caption{A function wrapped with a contract stating that it always returns
% an even number}
% \label{fig:num-to-even}
% \end{figure}

@@

The soundness of all these optimizations is tied to the invariance of the
semantics of programs with respect to specific substitutions. Inlining requires
that one can replace the application of a function by its body, which is
basically $\beta$-reduction: as long as the arguments are evaluated following
the language's strategy, it is a valid transformation. On the other hand,
a CSE-like transformation on a term $M$ requires an expansion property of the form

\begin{equation}\label{eq:cbn-expansion}
M[x/N] \simeq let~x~=~N~in~M
\end{equation}

Here, $M[x/N]$ is the substitution of $x$ for $N$ in the term $M$.
\ref{eq:cbn-expansion} has an evident call-by-name flavour, and consequently
fails in general for a call-by-value language with side effects.  Indeed, define
the following function $f$:

\begin{lstlisting}[language=Nickel]
let f x = print "hi";(x+1) in 
\end{lstlisting}

Then:

\begin{lstlisting}[language=Nickel]
(f 1,f 1) (*@ $\not \simeq$ @*) let y = f 1 in (y,y)
\end{lstlisting}

In the example above, the first term prints \nickel{"hi"} two times while the
second term prints it only once. However, \ref{eq:cbn-expansion} does hold for
\emph{pure} terms, that are terms without side-effects. In a pure language, or a
language with effect tracking like Haskell\todo{Cite a different language, or
remove citation}, pure terms can be safely identified for applying
optimizations.

Strikingly, we will see in section~\ref{sec:issues-sem} that the introduction of
unions and intersections can make \ref{eq:cbn-expansion} \resolved{it} unsound
\emph{even in a pure setting}, making the optimization of
Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.

@@

\change{These next few labels need to be inside figures,
they are not working}
\label{fig:union-adt-ex}
\begin{lstlisting}[title={Tree-like data structure}]
type BinaryTree =
Integer @| (BinaryTree, BinaryTree)
\end{lstlisting}

In addition to internal program representations, unions are also useful to model
external data:
\begin{itemize}
    \item To represent nullable values (\ref{fig:union-nullable-ex}).
    \item For plain enumerations (\ref{fig:union-enum-ex}).
    \item To support different representations of the same data
        (\ref{fig:union-alt-ex}).
\end{itemize}
All of which are common in configurations.

\label{fig:union-nullable-ex}
\begin{lstlisting}[title={Nullable values}]
optionalValue : Str @| Null
\end{lstlisting}

\label{fig:union-enum-ex}
\begin{lstlisting}[title={Plain enumeration}]
protocol : `Http @| `Ftp @| `Sftp
\end{lstlisting}

\label{fig:union-alt-ex}
\begin{lstlisting}[title={Alternative representations}]
person : {name : Str, age : Int}
  @| {name : Str, dateOfBirth : Str}
\end{lstlisting}

@@

\label{fig:intersection-subytping-ex}
\begin{lstlisting}[title={Value subtyping}]
1 : Int @& Float
\end{lstlisting}

@@

\label{fig:intersection-overloading-ex}
\begin{lstlisting}[title={Overloading}]
incr: (Int -> Int) @& (Float -> Float) =
  fun x => x + 1
\end{lstlisting}

@@

\label{fig:intersection-mixins-ex}
\begin{lstlisting}[language=Scala,title={Structural mixins}]
Comparator: {cmp: T -> T -> Bool}
Shower: {show: T -> String}

@@

\label{fig:intersection-dual}
\begin{lstlisting}[title={Intersection as the dual of unions}]
f : (Num @| Str) -> Bool
(*@ $\simeq$ @*) f : (Num -> Bool) @& (Str -> Bool)
\end{lstlisting}

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f | Positive -> Positive
      = fun x => x - 7 in
(f 10) + (f 5)
\end{lstlisting}
\caption{Wrong positive to positive function}
\label{fig:pos-to-pos}
\end{figure}

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f | (Positive -> Positive)
        @| (Positive -> NonPositive)
      = fun x => x - 7 in
(f 10) + (f 5)
\end{lstlisting}
\caption{Still wrong positive to positive or positive to non-positive function}
\label{fig:wrong-union-function}
\end{figure}

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Inlined]
let f | Num -> (Bool -> Num @| Bool -> Str)
      = fun x y => if y then x else "False"
in (f 1 true, f 1 false)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Common Subexpression Eliminated]
let f = (fun x y => if y then x else "False") | Num -> (Bool -> Num @| Bool -> Str) in
let g = f 1 in
(g true, g false)
\end{lstlisting}
\caption{Equivalent programs, with inlining or CSE applied}
\label{fig:optimized-programs}
\end{figure}

Unexpectedly, these two programs behave very differently, the first one will
return \nickel{(1, "False")}, without failing.  While the second one will fail,
since the single application of \nickel{f 1}, applied to both \nickel{true} and
\nickel{false} will be able to find the error on the union contract
\nickel{(Bool -> Num) @| (Bool -> Str)}.

@@

Summing it up, wrapping a value with a union of higher order contracts means that
that value now has a state, that may get modified silently at every use site.
Both reviewed papers suffer from this defect since it is the semantics
of unions that interact badly with code manipulation operations,
not a particular implementation.
This addition of global state complicates both the work of the language
implementor, as well as the language user, since it adds non trivial
side effects to your code.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Implementation]
x | A @& B (*@ $\simeq$ @*) (x | A) | B
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Problem]
let g | Num -> Num @& Str -> Str
      = fun x => x in
g 1
\end{lstlisting}
\caption{Implementations of intersection contracts: first iteration}
\label{fig:inter-contracts}
\end{figure}

Not all hope is lost, we could pass some extra shared state
to each one of the sub contracts and let them only fail if the other
side of the contract has failed as well, as shown in Figure
\ref{fig:inter-contracts-2}, where the shared state is
represented by a label \nickel{l}.
This idea is still quite reasonable, for instance, when executing
\nickel{g 1}, from Figure \ref{fig:inter-contracts}, the first contract
that gets checked is the argument of \nickel{Str -> Str} over the function parameter, so
something like \nickel{1 | Str}.
Of course, this fails, but before raising a violation, this contract could
check if the other side of the intersection (\nickel{Number -> Number}) has
also tried to raise negative blame on the same parameter ()\nickel{1}), since
it hasn't, the contract would return normally, and would mark itself as failed.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Implementation]
x | A @& B (*@ $\simeq$ @*) (x | A[l]) | B[l]
\end{lstlisting}
\caption{Implementations of intersection contracts: second iteration}
\label{fig:inter-contracts-2}
\end{figure}

On the other hand, if you were to do something like
\nickel{g true}, that would fail also negatively the contract
\nickel{Str -> Str} and, when checking if \nickel{true}
is a correct parameter for \nickel{Number -> Number} it would check
if the other side of the intersection has failed, and since it
has, a violation would be raised and the whole program would
correctly fail.

This idea, which is an oversimplification of the ideas
presented by Williams, Morris, and Wadler, seems to be spot on.
However, it presents a major problem, it's not compatible with flat contracts,
as presented in Section \ref{sec:flat-contracts}.
Consider the example presented on Figure \ref{fig:inter-flat-contracts}, if the
implementation discussed previously were to be used, we may end up with
something like this \nickel{((fun x => x) | (Str -> Str)[l]) | C[l]}, where
\nickel{l} represents the shared state.
At this point, the intuitive thing to do is to check the contract \nickel{C}, by
checking if \nickel{((fun x => x) | (Str -> Str)[l]) 0 == 0}.  This will fail,
since applying a function wrapped in \nickel{Str -> Str} contract to \nickel{0}
fails negatively.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let C = fun f => (f 0) == 0 in
let g | (String -> String) @& C
      = (fun x => x) 
\end{lstlisting}
\caption{Intersection and flat contracts}
\label{fig:inter-flat-contracts}
\end{figure}

@@

These expectations are not really met by how unions and intersections
contracts are supposed to work.
For example, consider the Nickel code shown on Figure
\ref{fig:intersection-distrbution}, at first sight, this code should
complete execution successfully, returning \nickel{(1, 1)},
however, under both the Keil and Thiemann, as well as the
Williams, Morris, and Wadler works, this program will fail, with a violation
of the contract imposed to \nickel{g}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = fun x y => x in
let g = f | (Num -> Num -> Num)
            @& (Num -> Bool -> Num) in
let h = g 1 in
(h 1, h true)
\end{lstlisting}
\caption{Intersection contracts don't distribute}
\label{fig:intersection-distrbution}
\end{figure}

The reason this happens is that there is a fundamental property of how
intersection contracts behave: the choice of which branch to choose is made
exactly when the intersection is opened up, and not delayed as much as possible,
as one would like.  This means, that the contract presented on Figure
\ref{fig:intersection-distrbution}, \nickel{Num -> Num -> Num @& Num
-> Bool -> Num}, will not behave the same way that \nickel{Num ->
(Num -> Num @& Bool -> Num)}, when, with an intuitive view of how
intersections should behave, these two expressions should be logically
equivalent.

It is worth noting, that intersection types do have this property,
since subtyping would allow to correctly type check a term like
\nickel{g 1} as both \nickel{Number -> Number} as well as
\nickel{Bool -> Num}, and then deducing that \nickel{g 1}
can be given the type \nickel{Num -> Num @& Bool -> Num}.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
x | A @& (B @| C)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Transformed]
x | (A @& B) | (A @& C)
\end{lstlisting}
\caption{Contract transformation performed by the Keil and
Thiemann work}
\label{fig:kt-contract-transf}
\end{figure}

This straightforward idea makes composing the Keil and Thiemann
solution with other existing systems more difficult; in the words
of Williams, Morris, and Wadler (see Section~\ref{sec:will-morr-wadl}):
``(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
special cases can hinder composition.''\cite{RootCauseOfBlame}.
\info{(Yann) Does
it also make it less efficient? If $(A \cap B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
And it also could harm performance, or at least complicate an implementation,
since some sharing mechanism should be used to avoid checking the same contract,
over the same value, multiple times, exemplified by contract \nickel{A} on
Figure \ref{fig:kt-contract-transf}, that gets duplicated.
\unsure{Arnaud says: this whole
section is too abstract give concrete code and explain what goes
wrong. It's fine to give code in Nickel syntax}

@@

As an example, consider the programs shown on Figure
\ref{fig:kt-non-deterministic}, if the Original program
gets executed, it will spawn two instances of said program,
one with the first branch of the intersection (\nickel{3 | Num}),
and the other one with the second branch (\nickel{3 | Positive}),
and these will get joined once they reach an equal point, in this case
it will be \nickel{3}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
3 | Num @& Positive
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=First Program]
3 | Num
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Second Program]
3 | Positive
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Joined Program]
3
\end{lstlisting}
\caption{Non deterministic intersection contract opening}
\label{fig:kt-non-deterministic}
\end{figure}

The second set of rules is deterministic, and it is intended
to be simpler to implement.
However, this second set needs to perform operations
on the evaluation context, to decide whether or not to evaluate
a given contract.
\footnote{You can check more on this topic on Section 5.1 of
\cite{KeilThiemannUnionIntersection}}
For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
when using the value \nickel{(fun x => x) | Str -> Str)}
inside the execution of contract \nickel{C}, and applying it to \nickel{0},
the contract runtime would check the context of execution and it would
deduce that contract \nickel{Str -> Str does not apply in this case,
skipping it.

@@

Altough this is true, it's a weaker statement of what we'd like to have,
consider the two contexts presented on Figure \ref{fig:valid-contexts},
the first one is a valid context for a \nickel{Num -> Num} function,
that is, you could replace the $\square$ by some function wrapped in a
\nickel{Num -> Num} contract, and the context will not violate it.
Similarly, the second context from the same figure is a valid
\nickel{Bool -> Bool} context.
